---
title: 'Unit 4 Lecture 4: Boosting'
date: "October 30, 2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, echo = FALSE}
# set ggplot theme
ggplot2::theme_set(ggplot2::theme_bw())
options(warn=-1)
```

Today, we will learn how to train and tune boosting models using the `gbm` package. Note that the `xgboost` package is more popular in practice, but the `gbm` package is simpler to use and is a stepping stone to learning `xgboost`.

First, let's load some libraries:
```{r, message = FALSE}
library(gbm)
library(tidyverse)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, fig.align = "center")
```


# Boosting models for regression

Like last time, we will be using the `hitters` data, splitting into training and testing:
```{r, message = FALSE}
hitters_data <- read_csv("hitters-data.csv")

# gbm expects character features to be coded as factors
hitters_data <- hitters_data |>
  mutate(across(where(is.character), as.factor))

set.seed(1) # set seed for reproducibility
train_samples <- sample(1:nrow(hitters_data), round(0.8 * nrow(hitters_data)))
hitters_train <- hitters_data |> filter(row_number() %in% train_samples)
hitters_test <- hitters_data |> filter(!(row_number() %in% train_samples))
```

## Training a gradient boosting model

Arguments:

- `distribution`: "gaussian" for continuous responses; "bernoulli" for binary responses
- `n.trees`: maximum number of trees to try; defaults to 100 but this is normally not enough trees
- `interaction.depth`: interaction depth; defaults to 1
- `shrinkage`: shrinkage parameter lambda: defaults to 0.1
- `bag.fraction`: subsampling fraction pi; defaults to 0.5
- `cv.folds`: number of CV folds to use; defaults to 0 (i.e. no CV) but we prefer 5
- `train.fraction`: fraction of data to use as training; rest used as validation set; we leave at default of 1
- `n.cores`: how many parallel processors to use for CV; we set to 1

```{r}
# read more about the inputs and outputs, bells and whistles of gbm
?gbm
```

Training the model:
```{r}
set.seed(1)
gbm_fit <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
```

We can visualize the CV error using `gbm.perf`, which both makes a plot and outputs the optimal number of trees:
```{r}
opt_num_trees <- gbm.perf(gbm_fit)
opt_num_trees
```

The green curve is the CV error; the black curve is the training error. The dashed blue line indicates the minimum of the CV error. 

Note that `gbm_fit$cv.error` also contains the CV errors, so these can be plotted manually as well:
```{r}
ntrees <- 100
tibble(Iteration = 1:ntrees, CV = gbm_fit$cv.error) |>
  ggplot(aes(x = Iteration, y = CV)) +
  geom_line()
```

We want to make sure there are enough trees that the CV curve has reached its minimum. For example, suppose we had chosen a smaller shrinkage parameter, e.g. 0.01:

```{r}
set.seed(1)
gbm_fit_slow <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.01,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
gbm.perf(gbm_fit_slow)
```
We see that 100 is not enough trees for lambda = 0.01. In this case, we would need to increase the number of trees:

```{r}
set.seed(1)
gbm_fit_slow <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 1000,
  interaction.depth = 1,
  shrinkage = 0.01,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
gbm.perf(gbm_fit_slow)
```

## Tuning the interaction depth

The quick way to tune the interaction depth is to try out a few different values:
```{r}
set.seed(1)
gbm_fit_1 <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
gbm_fit_2 <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 2,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
gbm_fit_3 <- gbm(Salary ~ .,
  distribution = "gaussian",
  n.trees = 100,
  interaction.depth = 3,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = hitters_train
)
```

We can extract the CV errors from each of these objects by using the `cv.error` field:
```{r}
ntrees <- 100
cv_errors <- bind_rows(
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_1$cv.error, depth = 1),
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_2$cv.error, depth = 2),
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_3$cv.error, depth = 3)
)
cv_errors
```

We can then plot these as follows:
```{r}
cv_errors |>
  ggplot(aes(x = ntree, y = cv_err, colour = factor(depth))) +
  geom_line()
```

Which value of `interaction.depth` seems to work the best here?

Let's save the optimal model and optimal number of trees (note `plot.it = FALSE` in `gbm.perf` to extract the optimal number of trees without making the CV plot again):
```{r}
gbm_fit_optimal <- gbm_fit_3
optimal_num_trees <- gbm.perf(gbm_fit_3, plot.it = FALSE)
optimal_num_trees
```

## Model interpretation

Let's now interpret our tuned model. To get the variable importance measures, we use `summary`, specifying the number of trees via the `n.trees` argument:
```{r}
summary(gbm_fit_optimal, n.trees = optimal_num_trees, plotit = FALSE)
```

We can also make the partial dependence plots for the different features using `plot`:
```{r}
plot(gbm_fit_optimal, i.var = "CAtBat", n.trees = optimal_num_trees)
plot(gbm_fit_optimal, i.var = "CRuns", n.trees = optimal_num_trees)
```

## Making predictions based on a boosting model:

We can make predictions using `predict`, as usual, but we need to specify the number of trees to use:
```{r}
gbm_predictions <- predict(gbm_fit_optimal,
  n.trees = optimal_num_trees,
  newdata = hitters_test
)
gbm_predictions
```
We can compute the mean-squared prediction error as usual too:
```{r}
mean((gbm_predictions - hitters_test$Salary)^2)
```

# Boosting for classification

Boosting models work very similarly for classification. Let's continue with the heart disease data from last time: 
```{r, message = FALSE, warning = FALSE}
heart_data <- read_csv("heart-data.csv")
heart_data <- heart_data |>
  na.omit() |>
  mutate(AHD = ifelse(AHD == "Yes", 1, 0)) |> # gbm expects response to be 0-1,
  #  NOT factor (unlike RF)
  mutate_at(c("ChestPain", "Thal"), as.factor)

set.seed(1) # set seed for reproducibility
train_samples <- sample(1:nrow(heart_data), round(0.8 * nrow(heart_data)))
heart_train <- heart_data |> filter(row_number() %in% train_samples)
heart_test <- heart_data |> filter(!(row_number() %in% train_samples))
```

Fitting a boosting model uses the same basic syntax, but with `distribution = "bernoulli"`:
```{r}
set.seed(1)
gbm_fit <- gbm(AHD ~ .,
  distribution = "bernoulli",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = heart_train
)
```

Common pitfalls when fitting a `gbm`:

- The binary response is coded as a `character`, e.g. "Yes"/"No".
- The binary response is coded as a `factor`. 
- Any of the features are coded as strings, rather than factors.

```{r}
gbm.perf(gbm_fit)
```

We can tune the interaction depth in the same way as before:

```{r}
# try a few values
set.seed(1)
gbm_fit_1 <- gbm(AHD ~ .,
  distribution = "bernoulli",
  n.trees = 100,
  interaction.depth = 1,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = heart_train
)
set.seed(1)
gbm_fit_2 <- gbm(AHD ~ .,
  distribution = "bernoulli",
  n.trees = 100,
  interaction.depth = 2,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = heart_train
)
set.seed(1)
gbm_fit_3 <- gbm(AHD ~ .,
  distribution = "bernoulli",
  n.trees = 100,
  interaction.depth = 3,
  shrinkage = 0.1,
  cv.folds = 5,
  n.cores = 1,
  data = heart_train
)

# extract CV errors
ntrees <- 100
cv_errors <- bind_rows(
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_1$cv.error, depth = 1),
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_2$cv.error, depth = 2),
  tibble(ntree = 1:ntrees, cv_err = gbm_fit_3$cv.error, depth = 3)
)

# plot CV errors
cv_errors |>
  ggplot(aes(x = ntree, y = cv_err, colour = factor(depth))) +
  geom_line()
```

Aha! We see some overfitting! For which values of interaction depth do we see more overfitting, and why? What is the optimal interaction depth?

```{r}
gbm_fit_optimal <- gbm_fit_1
optimal_num_trees <- gbm.perf(gbm_fit_1, plot.it = FALSE)
```

We can calculate variable importance scores as before:
```{r}
summary(gbm_fit_optimal, n.trees = optimal_num_trees, plotit = FALSE)
```

For the partial dependence plots, it's useful to specify `type = "response"` so we can interpret the y axis on the probability scale:
```{r}
plot(gbm_fit_optimal, 
     i.var = "ChestPain", 
     n.trees = optimal_num_trees, 
     type = "response")

plot(gbm_fit_optimal, 
     i.var = "Chol", 
     n.trees = optimal_num_trees, 
     type = "response")
```

To make predictions, use the same syntax as before but with `type = "response"` to get predictions on the probability scale:
```{r}
gbm_probabilities <- predict(gbm_fit_optimal,
  n.trees = optimal_num_trees,
  type = "response", newdata = heart_test
)
gbm_probabilities
```

We can then threshold the probabilities at 0.5 as usual and calculate the misclassification error:
```{r}
gbm_predictions <- as.numeric(gbm_probabilities > 0.5)
mean(gbm_predictions != heart_test$AHD)
```
