---
title: 'Unit 4 Lecture 2: Pruning and cross-validating decision trees'
date: "October 23, 2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# options(scipen = 0, digits = 3) # controls number of significant digits printed
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot theme
options(warn=-1) # optional, to suppress warnings; comment out if you want to see warnings

#' cv_tree: Make a CV plot and pick tree sizes (min and 1-SE) from an rpart fit
#'
#' @param fit an rpart model object (classification or regression)
#' @return a list with:
#'   - df: data.frame of CP table augmented with nleaves
#'   - cv_plot: ggplot object (CV error vs number of terminal nodes)
#'   - nleaves.min, nleaves.1se: chosen sizes (terminal nodes)
#'   - CP.min, CP.1se: corresponding complexity parameters
#'   - xerror.min, xerror.1se: CV error at the chosen sizes
#'
cv_tree <- function(fit) {
  if (!inherits(fit, "rpart")) stop("fit must be an rpart object")

  if (is.null(fit$cptable)) {
    stop("No cptable found. Fit with rpart and ensure xval > 1 (default is 10).")
  }

  # Extract CP table and augment
  tab <- as.data.frame(fit$cptable)
  # rpart names vary slightly; standardize
  names(tab) <- sub("^CP$", "CP", names(tab))
  if (!all(c("CP", "nsplit", "rel error", "xerror", "xstd") %in% names(tab))) {
    stop("Unexpected cptable columns; got: ", paste(names(tab), collapse=", "))
  }

  tab$nleaves <- tab$nsplit + 1L

  # Min-CV pick
  i_min <- which.min(tab$`xerror`)
  xerr_min <- tab$`xerror`[i_min]
  se_min   <- tab$`xstd`[i_min]
  thr_1se  <- xerr_min + se_min

  # 1-SE pick: simplest (fewest leaves) whose CV error <= min + 1*SE
  idx_ok <- which(tab$`xerror` <= thr_1se)
  i_1se  <- idx_ok[which.min(tab$nleaves[idx_ok])]

  out <- list(
    df = tab[, c("CP", "nsplit", "nleaves", "rel error", "xerror", "xstd")]
  )

  out$nleaves.min <- tab$nleaves[i_min]
  out$nleaves.1se <- tab$nleaves[i_1se]
  out$CP.min      <- tab$CP[i_min]
  out$CP.1se      <- tab$CP[i_1se]
  out$xerror.min  <- tab$`xerror`[i_min]
  out$xerror.1se  <- tab$`xerror`[i_1se]

  # Build plot (style matching the handout)
  if (!requireNamespace("ggplot2", quietly = TRUE)) {
    warning("Install ggplot2 to get the plot: install.packages('ggplot2')")
    out$cv_plot <- NULL
    return(out)
  }
  library(ggplot2)

  p <- ggplot(out$df, aes(x = nleaves, y = `xerror`)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = `xerror` - `xstd`, ymax = `xerror` + `xstd`), width = 0.2) +
    geom_hline(yintercept = xerr_min, linetype = "dashed") +
    geom_hline(yintercept = thr_1se, linetype = "dashed") +
    geom_vline(xintercept = out$nleaves.min, linetype = "dashed") +
    geom_vline(xintercept = out$nleaves.1se, linetype = "dashed") +
    labs(x = "Number of terminal nodes", y = "CV error") +
    theme_minimal(base_size = 12)

  out$cv_plot <- p
  out
}
```

Today, we will learn how to select the complexity of decision trees based on cost complexity pruning and cross-validation, as implemented in the `rpart` package.

First, let's load some libraries:
```{r, message = FALSE}
library(rpart)       # for training decision trees
library(rpart.plot)  # for plotting decision trees 
library(tidyverse)   # for everything else
```

# Regression trees

Like last time, we will be using the `hitters` data, splitting into training and testing:
```{r, message = FALSE}
hitters_data <- read_csv("hitters-data.csv")
set.seed(1) # set seed for reproducibility
train_samples <- sample(1:nrow(hitters_data), round(0.8 * nrow(hitters_data)))
hitters_train <- hitters_data |> filter(row_number() %in% train_samples)
hitters_test <- hitters_data |> filter(!(row_number() %in% train_samples))
```

As before, we fit a regression tree by calling `rpart`:
```{r}
tree_fit <- rpart(Salary ~ ., data = hitters_train)
```

## Tree pruning and cross validation

It turns out that in addition to growing the tree, behind the scenes `rpart` has already:

- used cost complexity pruning to get the nested sequence of trees
- applied 10-fold cross-validation to compute the CV estimates and standard errors for each value of $\alpha$

All we need to do is call the `printcp` function to get a summary of all this information:
```{r}
printcp(tree_fit)
```
Let's focus on the table at the bottom of this output. Each row corresponds to a tree in the sequence obtained by pruning. Let's discuss each column in turn:

- The `CP` column is the "complexity parameter". It is related to, but not exactly the same as, the $\alpha$ parameter from the slides. Be careful! The terminology "complexity parameter" is a bit misleading because higher complexity parameters correspond to less complex models (just like lambda in penalized regression). 
- `nsplit` is the number of splits in the tree. Note that `1+nsplit` is the number of terminal nodes in the tree.
- `rel error` is the RSS training error of the tree, normalized by the total variance of the response; equivalently, this is $1-R^2$. The training error decreases as the complexity increases.
- `xerror` is the cross-validation error estimate.
- `xstd` is the cross-validation standard error.

The exact values of the complexity parameter are not so important; we might as well parameterize the trees based on the number of terminal nodes. Armed with all this information, we can produce a CV plot using `cv_tree()` from `stat471`: 
```{r, fig.width = 4, fig.height = 3}
tree_cv_info <- cv_tree(tree_fit)
tree_cv_info$cv_plot
```

The `tree_cv_info` object also has additional information, like `nleaves.1se` and `nleaves.min`. There are also fields called `CP.1se` and `CP.min`, containing the corresponding CP values. These will be useful below. 

## Extracting the pruned tree and making predictions

To actually get the optimal pruned tree, we need to use the function `prune`, specifying the complexity parameter CP. We can get this complexity parameter from the `tree_cv_info` object:
```{r}
optimal_tree <- prune(tree_fit, cp = tree_cv_info$CP.1se)
```

As before, we can plot this tree using `rpart.plot`:
```{r, fig.width = 4, fig.height = 3}
rpart.plot(optimal_tree)
```

That is a small tree! In the bias variance trade-off, sometimes less (complexity) is more (predictive performance). 

Now we can make predictions on the test data and evaluate MSE using this tree:
```{r}
pred <- predict(optimal_tree, newdata = hitters_test)
pred

mean((pred - hitters_test$Salary)^2)
```

# Exercise: Classification trees

Let's continue with the heart disease data from last time: 
```{r, message = FALSE, warning = FALSE}
heart_data <- read_csv("heart-data.csv")
set.seed(1) # set seed for reproducibility
train_samples <- sample(1:nrow(heart_data), round(0.8 * nrow(heart_data)))
heart_train <- heart_data |> filter(row_number() %in% train_samples)
heart_test <- heart_data |> filter(!(row_number() %in% train_samples))

# fit a classification tree
tree_fit <- rpart(AHD ~ .,
  method = "class",               # classification
  parms = list(split = "gini"),   # Gini index for splitting
  data = heart_train
)
```

## Tree pruning and cross-validation

1. Produce the table of the trees in the sequence obtained from cost complexity pruning. How does `nsplit` vary with `CP`? Does this relationship make sense?
 

2. Produce the CV plot. How many terminal nodes would we choose based on the one-standard-error rule? Do we notice anything strange about the CV plot?
 

3. Extract and visualize the tree chosen by cross-validation. How many terminal nodes does the tree have?
 

    
4. What is the test misclassification error of this decision rule? 
 