---
title: 'STAT 4710/5710: Homework 2'
author: 'David Dai'
date: 'Sep 25, 2025'
output:
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: '3'
urlcolor: blue
---

```{r setup, include=FALSE}
options(scipen = 0, digits = 3) # controls number of significant digits printed
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot theme
options(warn=-1) # optional, to suppress warnings; comment out if you want to see warnings
```

```{r, echo=FALSE}
cross_validate_spline <- function(x, y, nfolds = 10, df_values = 2:20, seed = NULL) {
  stopifnot(length(x) == length(y))
  n <- length(x)
  if (!is.null(seed)) set.seed(seed)

  # Same fold construction style as in the Rmd
  folds <- sample(rep(1:nfolds, length.out = n))
  train_df <- data.frame(x = x, y = y, fold = folds)

  # Helper: compute CV mean & SE for a single displayed df value
  # NOTE: to match your validation plot, we fit ns(x, df = df_display - 1)
  cv_for_df <- function(df_display) {
    df_use <- df_display - 1
    oof <- numeric(n)
    for (k in 1:nfolds) {
      out <- train_df[train_df$fold != k, , drop = FALSE]
      in_  <- train_df[train_df$fold == k, , drop = FALSE]
      fit <- lm(y ~ ns(x, df = df_use), data = out)
      oof[folds == k] <- predict(fit, newdata = in_)
    }
    # fold-wise CV estimates (same aggregation as in the Rmd)
    cv_by_fold <- tapply((oof - train_df$y)^2, INDEX = train_df$fold, FUN = mean)
    c(mean = mean(cv_by_fold), se = sd(cv_by_fold) / sqrt(nfolds))
  }

  # Run across all df values
  mat <- t(vapply(df_values, cv_for_df, numeric(2)))
  cv_table <- data.frame(
    df = df_values,
    cv_mean = mat[, "mean"],
    cv_se   = mat[, "se"]
  )

  # df.min and 1SE rule (prefer simpler/smaller df within 1 SE)
  min_row <- cv_table[which.min(cv_table$cv_mean), ]
  df.min  <- min_row$df
  thresh  <- min_row$cv_mean + min_row$cv_se
  df.1se  <- with(cv_table[cv_table$cv_mean <= thresh, ], min(df))

  # Plot (points + line + error bars; dashed lines at min and 1SE)
  cv_plot <- ggplot(cv_table, aes(df, cv_mean)) +
    geom_line(linewidth = 0.8, color = "black") +
    geom_point(size = 2, color = "black") +
    geom_errorbar(aes(ymin = cv_mean - cv_se, ymax = cv_mean + cv_se),
                  width = 0.3, color = "black") +
    geom_hline(yintercept = min_row$cv_mean, linetype = 2) +
    geom_hline(yintercept = thresh, linetype = 2) +
    geom_vline(xintercept = df.min, linetype = 3) +
    geom_vline(xintercept = df.1se, linetype = 3) +
    labs(x = "Degrees of freedom", y = "CV error") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  list(cv_table = cv_table, cv_plot = cv_plot, df.1se = df.1se, df.min = df.min)
}
```

# Instructions {.unnumbered}

## Materials and collaboration {.unlisted .unnumbered}

The policy on allowed materials and collaboration is as stated on the Syllabus:

>"Students are permitted to work together on homework assignments, but must write up and submit solutions individually. In particular, students may not copy each others’ solutions. Students may consult all course materials, textbooks, the internet, or AI tools (e.g. ChatGPT) to complete their homework. Students may not use solutions to problems that may be available online and/or from past iterations of the course. For each homework, students must disclose all classmates with whom they collaborated, which AI tools they used, and how they used them. Failure to do so will result in a 10-point penalty."

In accordance with this policy, 

*Please disclose all classmates with whom you collaborated:* 
NA

*Please disclose which AI tools you used, and how you used them:* 
I used ChatGPT for R code grammar example

\textbf{\textcolor{red}{Failure to answer the above questions will result in a 10-point penalty.}}

## Writeup {.unlisted .unnumbered}

\textcolor{blue}{Use} `homework-2-template.Rmd` \textcolor{blue}{as a starting point for your writeup, adding your solutions after "**Solution**". Add your R code using code chunks and add your text answers using **bold text**.} Consult the [preparing reports guide](https://katsevich-teaching.github.io/stat-4710-fall-2023/assets/preparing-reports.pdf) for guidance on compilation, creation of figures and tables, and presentation quality. If the instructions ask you to "print a tibble", you should print the tibble directly. 


## Programming {.unlisted .unnumbered}

We encourage you to follow the `tidyverse` paradigm for data visualization, manipulation, and wrangling, as in the first homework, to practice tidyverse workflows (e.g., `filter`, `mutate`, `group_by`, `summarize`, `join`, `ggplot2`).

Students who strongly prefer to use Python for homework may do so. If you choose this option, you are **required** to present both code chunks and the corresponding outputs/interpretations in your PDF. In addition, it's recommended that you:

- Use `pandas` for data manipulation (preferred, since it parallels `dplyr`).
- Use `plotnine` for visualization (preferred, since it parallels ggplot2). If using seaborn/matplotlib, you must still include the required plot elements (labels, 45° line, smoothing curve, facets, etc.).  
 

```{r, message=FALSE}
library(tidyverse)
library(splines) 
```

**Code completion.** In some problems, the task is to complete an incomplete code chunk provided. In those problems, in the template, there is `eval = FALSE` in the chunk header to prevent evaluation errors. You should remove `eval = FALSE` after you have completed the code to enable evaluation and printing of results.

## Grading {.unlisted .unnumbered}
The point value for each problem sub-part is indicated. There are 100 points possible on this homework, evaluated based on the correctness of the numerical results and plots. When a problem asks for a plot:

- **Axes and labels**: Always include informative axis labels and, if appropriate, titles.
- **Required elements**: If the instructions ask for a 45° line, smoothing curve, or facets, these must be included.
- **Clarity**: Points may be deducted if plots are cluttered, unreadable, or missing labels.

## Submission {.unlisted .unnumbered}

Compile your writeup to PDF and submit to Gradescope.

\newpage

# Case study: spline regression model selection (64 points)
 

## Import data (2 points)

**Solution.** 
```{r}
library(tidyverse)
wage_df = read_csv("wages.csv")
# log transform wage column 
wage_df$wage = log(wage_df$wage)
print(wage_df)
```

## Tidy (2 points)

**Solution.** 
```{r}
# clean the wage_df
wage_df = wage_df |> 
          select(wage, age, education) |> 
          drop_na() |> 
          filter(education %in% c("HS Grad", "College Grad"))

print(wage_df)
```
 

## Explore (8 points)

**Solution.** 
```{r}
# check count of each education level
print(table(wage_df$education))

# Plot distribution of wage based on education level
ggplot(wage_df, aes(x = education, y=wage)) +
  geom_boxplot() + 
  labs(
    title = "Wage distribution by education",
    x = "Education level",
    y = "Wage"
  )

# plot relationship between age and wage
ggplot(wage_df, aes(x = age, y = wage)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~education) +
  labs(
    title = "Wage vs Age by Education Level",
    x = "Age",
    y = "Wage"
  )
```
**There are 971 HS Grad and 685 College Grad. For wages among different educational level, College Grad have a higher median wage compared with High School Grad; the wage variance among college grad is also smaller than those of high school grad, particularly in the age group between 40 to 60**

## Prepare for modeling (1 point)
 

```{r}
set.seed(4710) # set seed for reproducibility
n <- nrow(wage_df)
train_samples <- sample(1:n, size = 0.8 * n, replace = FALSE)
wage_train = wage_df[train_samples,]
wage_test = wage_df[-train_samples,]
```
 
## Validation set approach to model selection (33 points)


### Data splitting (3 points)

**Solution.** 
```{r}
# Split train_test based on education level
wage_train_HS = wage_train[wage_train$education == "HS Grad",]
wage_train_coll = wage_train[wage_train$education == "College Grad",]
wage_test_HS = wage_test[wage_test$education == "HS Grad",]
wage_test_coll = wage_test[wage_test$education == "College Grad",]

# Create learning and validation set
set.seed(4710) # set seed for reproducibility
HS_n <- nrow(wage_train_HS)
Coll_n = nrow(wage_train_coll)
HS_learn_samples <- sample(1:HS_n, size = 0.7 * HS_n, replace = FALSE)
coll_learn_samples <- sample(1:Coll_n, size = 0.7 * Coll_n, replace = FALSE)

wage_learn_HS =  wage_train_HS[HS_learn_samples,]
wage_val_HS = wage_train_HS[-HS_learn_samples,]
wage_learn_coll = wage_train_coll[coll_learn_samples,]
wage_val_coll = wage_train_coll[-coll_learn_samples,]
```
 
### Estimate MSE given a df (15 points)

**Solution.** 

```{r}
library(splines)
validation_mse <- function(learn_data, val_data, df) {
  spline_fit <- lm(wage ~ ns(age, df=df), data = learn_data)
  predictions <- predict(spline_fit, newdata = val_data)
  mse <- mean((predictions - val_data$wage)^2)
  return(mse)
}
df_seq = seq(2,20,1)
val_errors_HS = numeric(length(df_seq))
val_errors_coll = numeric(length(df_seq))
for (i in 1:length(df_seq)) {
  val_errors_HS[i] = validation_mse(wage_learn_HS, wage_val_HS, df_seq[i])
  val_errors_coll[i] = validation_mse(wage_learn_coll, wage_val_coll, df_seq[i])
}
print(val_errors_HS)
print(val_errors_coll)
```
 
### Model selection based on estimated MSE (10 points)

**Solution.** 
```{r}
# plots for HS grad degree of freedom vs. mse
mse_df <- data.frame(
  df   = rep(df_seq, 2),
  mse  = c(val_errors_HS, val_errors_coll),
  group = rep(c("HS Grad", "College Grad"), each = length(df_seq))
)

ggplot(mse_df, aes(x = df, y = mse, color = group)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Validation MSE vs Degrees of Freedom",
    x = "Degrees of Freedom (df)",
    y = "Validation MSE",
    color = "Education Group"
  ) +
  theme_minimal()
```
**For HS Grad group, df=14 minimizes the validation MSE; for College Grad, df=4 minimizes the validation MSE.**

### Fit chosen model and assessment (5 points)

**Solution.** 
```{r}
df_HS = 14
df_coll = 4
HS_spline_fit = lm(wage~ns(age, df = df_HS), data = wage_train_HS)
coll_spline_fit = lm(wage~ns(age, df = df_coll), data = wage_train_coll)
HS_spline_predict = predict(HS_spline_fit, newdata=wage_test_HS)
coll_spline_predict = predict(coll_spline_fit, newdata=wage_test_coll)
HS_mse = mean((HS_spline_predict - wage_test_HS$wage)^2)
coll_mse = mean((coll_spline_predict - wage_test_coll$wage)^2)
print(HS_mse)
print(coll_mse)
```

**Spline with best df from validation set approach for HS Grad is 0.00479 and for College Grad is 0.00486**
 
## Cross validation model selection (18 points)

### Model selection (11 points)

```{r, echo = FALSE}
cross_validate_spline <- function(x, y, nfolds = 10, df_values = 2:15, seed = NULL) {
  stopifnot(length(x) == length(y))
  n <- length(x)
  if (!is.null(seed)) set.seed(seed)

  # Same fold construction style as in the Rmd
  folds <- sample(rep(1:nfolds, length.out = n))
  train_df <- data.frame(x = x, y = y, fold = folds)

  # Helper: compute CV mean & SE for a single displayed df value
  # NOTE: to match your validation plot, we fit ns(x, df = df_display - 1)
  cv_for_df <- function(df_display) {
    df_use <- df_display - 1
    oof <- numeric(n)
    for (k in 1:nfolds) {
      out <- train_df[train_df$fold != k, , drop = FALSE]
      in_  <- train_df[train_df$fold == k, , drop = FALSE]
      fit <- lm(y ~ ns(x, df = df_use), data = out)
      oof[folds == k] <- predict(fit, newdata = in_)
    }
    # fold-wise CV estimates (same aggregation as in the Rmd)
    cv_by_fold <- tapply((oof - train_df$y)^2, INDEX = train_df$fold, FUN = mean)
    c(mean = mean(cv_by_fold), se = sd(cv_by_fold) / sqrt(nfolds))
  }

  # Run across all df values
  mat <- t(vapply(df_values, cv_for_df, numeric(2)))
  cv_table <- data.frame(
    df = df_values,
    cv_mean = mat[, "mean"],
    cv_se   = mat[, "se"]
  )

  # df.min and 1SE rule (prefer simpler/smaller df within 1 SE)
  min_row <- cv_table[which.min(cv_table$cv_mean), ]
  df.min  <- min_row$df
  thresh  <- min_row$cv_mean + min_row$cv_se
  df.1se  <- with(cv_table[cv_table$cv_mean <= thresh, ], min(df))

  # Plot (points + line + error bars; dashed lines at min and 1SE)
  cv_plot <- ggplot(cv_table, aes(df, cv_mean)) +
    geom_line(linewidth = 0.8, color = "black") +
    geom_point(size = 2, color = "black") +
    geom_errorbar(aes(ymin = cv_mean - cv_se, ymax = cv_mean + cv_se),
                  width = 0.3, color = "black") +
    geom_hline(yintercept = min_row$cv_mean, linetype = 2) +
    geom_hline(yintercept = thresh, linetype = 2) +
    geom_vline(xintercept = df.min, linetype = 3) +
    geom_vline(xintercept = df.1se, linetype = 3) +
    labs(x = "Degrees of freedom", y = "CV error") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  list(cv_table = cv_table, cv_plot = cv_plot, df.1se = df.1se, df.min = df.min)
}
```
**Solution.** 
```{r}
cv_HS = cross_validate_spline(wage_train_HS$age,
                              wage_train_HS$wage,
                              nfolds = 10,
                              df_values = 2:20,
                              seed = 4710)
cv_coll = cross_validate_spline(wage_train_coll$age,
                              wage_train_coll$wage,
                              nfolds = 10,
                              df_values = 2:20,
                              seed = 4710)
print(cv_HS$cv_plot)
print(cv_coll$cv_plot)
```
**For HS Grad cv approach, the df.min is 7 and df.1se is 4; for Coll Grad cv approach, df.min is 9 and df.1se is 3. Base on the cv plot, I would choose df=4 for HS Grad spline and df=3 for college Grad spline. The validation set approach has resulting best df different from CV approach. Based on the MSE plots on both approaches, I would choose df=4 for HS Grad and df=3 for Coll Grad, since higher df does not provide significant better MSE on both approaches.**

### Final fit (5 points)

**Solution.** 
```{r}
df_hs_cv = 4
df_coll_cv = 3
HS_spline_cv = lm(wage~ns(age, df = df_hs_cv), data = wage_train_HS)
coll_spline_cv = lm(wage~ns(age, df = df_coll_cv), data = wage_train_coll)

# get testing mse
HS_spline_predict_cv = predict(HS_spline_cv, newdata=wage_test_HS)
coll_spline_predict_cv = predict(coll_spline_cv, newdata=wage_test_coll)
HS_mse_cv = mean((HS_spline_predict_cv - wage_test_HS$wage)^2)
coll_mse_cv = mean((coll_spline_predict_cv - wage_test_coll$wage)^2)
print(HS_mse_cv)
print(coll_mse_cv)

# also extract training mse
HS_train_preds <- predict(HS_spline_cv, newdata = wage_train_HS)
HS_train_mse <- mean((wage_train_HS$wage - HS_train_preds)^2)
coll_train_preds <- predict(coll_spline_cv, newdata = wage_train_coll)
coll_train_mse <- mean((wage_train_coll$wage - coll_train_preds)^2)
print(HS_train_mse)
print(coll_train_mse)
```
**I chose df.1se for both education levels. Training and testing MSE for HS_Grad is 0.00444 and 0.00452; training and testing MSE for Coll_Grad is 0.00495 and 0.00488. The training error is smaller than testing error for both HS Grad and Coll Grad; however, since training and testing error do not differ much, the occurred overfitting is not that substantial.**

### Compare with validation set approach (2 points)

**Solution.** 
**Validation set approach has test MSE 0.00479 for HS Grad and 0.00486 for College Grad. CV approach has test MSE 0.00452 for HS Grad and 0.00488 for College Grad. In HS Grad case, degree of freedom (df) chosen by CV is better than validation set approach; In College Grad case, df chosen by validation set is slightly better than CV.**

\newpage 
# Case study: kNN classification with Default data (36 points)
 

```{r}
library(ISLR2)
data("Default")
```

## Data split (2 points)

**Solution.** 
```{r}
set.seed(4710) # set seed for reproducibility
n_default <- nrow(Default)
train_samples_d <- sample(1:n_default, size = 0.8 * n_default, replace = FALSE)
default_train = Default[train_samples_d,]
default_test = Default[-train_samples_d,]
```

## Class imbalance and baseline (6 points)

**Solution.** 
```{r}
# check distribution of training dataset
print(table(default_train$default))

# check majority class
majority_class = names(which.max(table(default_train$default)))
print(majority_class)

# make baseline predictions
baseline_preds = rep(majority_class, nrow(default_test))

# check accuracy 
baseline_accuracy = mean(baseline_preds == default_test$default)
print(baseline_accuracy)
```
**The "No" class has 7740 data points, which is far more than "Yes" class, which has 260 data points in the training dataset. For the baseline classifer which always report "No", the accuracy on test dataset is 0.964. However, since in this case false negative have a huge impact on the prediction results, accuracy is not a good metric and a classifier only predict no is not a good one.**

## kNN classification (28 points)


### Load results 

**Solution.** 
```{r}
knn_counts = read_csv("knn_counts_by_K.csv")
```

### Compute error metrics in classification (4 points)

**Solution.** 
```{r}
# calculate accuracy, precision, recall and F1_score
knn_counts$accuracy = (knn_counts$TN + knn_counts$TP) / (knn_counts$TN + knn_counts$TP + knn_counts$FN + knn_counts$FP)
knn_counts$precision = knn_counts$TP / (knn_counts$TP + knn_counts$FP)
knn_counts$recall = knn_counts$TP / (knn_counts$TP + knn_counts$FN)
knn_counts$F1_score = 2 * (knn_counts$precision * knn_counts$recall) / (knn_counts$precision + knn_counts$recall)
print(knn_counts)
```

 
### Plot metrics vs. K (6 points)

**Solution.** 
```{r}
# get only learning and validation split 
knn_training = knn_counts[knn_counts$split %in% c("learning", "validation"),]

# plot based on accuracy
ggplot(data=knn_training, aes(x = K, y=accuracy, color = split)) +
  geom_line() + geom_point() + 
  labs(
    title = "K v.s. accuracy",
    x = "K",
    y = "accuracy",
    color = "split"
  )

# plot based on recall
ggplot(data=knn_training, aes(x = K, y=recall, color = split)) +
  geom_line() + geom_point() + 
  labs(
    title = "K v.s. recall",
    x = "K",
    y = "recall",
    color = "split"
  )
```
**I noticed the recall is much less than the accuracy level for both learning and validation data. The accuracy level is higher than 0.95 but recall level are below 0.4 for validation dataset. Based on recall formula TP / (TP + FN), one problem of the KNN approach is the large number of FN with respect to TP, resulting a low recall rate.**

### Model selection and improvement (conceptual) (8 points)
 
**Solution.** 
**As a risk manager, I would prefer recall as a metric than accuracy, since I am more concerned with the proportion of false negative that is of a high cost. Using recall and select a K based on validation approach is not a good idea, since we should also consider precision, which can inform when we have much false positives; otherwise, we may come up with a model classifies everything to "Yes". Also, the majority vote for KNN might be an issue since in default data, most of data are "No", only handful are "Yes" and the majority vote mechanism will make KNN more likely to classify "No" as an answer for default. To solve this issue, we can consider adding weights to "No" answer during training to combat the class imbalance problem.**

### Model complexity (conceptual) (10 points)

**Solution.** 
**The model complexity increases as K decreases, since a smaller K will make the model more sensitive to slightly change in neighborhood area, hence increase model complexity. The df n/K for KNN model makes sense because the model complexity is inverse proportional with K and proportional with n as more data in the training dataset will make the KNN model to get more diversified predictions based on training data points, which increases model complexity. When K is increased, the model is less likely be influenced by noise in training data, which decreases the variance of the model hence getting better predictions. However, increasing K too large may make the model oversimplified, which increase the bias of the model, hence leading to worse predictions.**
 