---
title: 'STAT 4710/5710: Homework 3'
author: 'David Dai' 
output:
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: '3'
urlcolor: blue
---

```{r setup, include=FALSE}
options(scipen = 0, digits = 3) # controls number of significant digits printed
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot theme
options(warn=-1) # optional, to suppress warnings; comment out if you want to see warnings
options(show.signif.stars = FALSE)
```

```{r, echo=FALSE}
coef_tidy <- function(fit, s = NULL) {
  # pick the underlying glmnet fit and lambda
  if (inherits(fit, "cv.glmnet")) {
    if (is.null(s)) s <- "lambda.1se"
    if (is.character(s)) {
      if (s == "lambda.min")  s_val <- fit$lambda.min
      else if (s == "lambda.1se") s_val <- fit$lambda.1se
      else stop("For cv.glmnet, s must be 'lambda.min', 'lambda.1se', or a numeric lambda.")
    } else if (is.numeric(s) && length(s) == 1) {
      s_val <- s
    } else stop("Invalid 's' for cv.glmnet.")
    co <- as.matrix(stats::coef(fit$glmnet.fit, s = s_val))
  } else if (inherits(fit, "glmnet")) {
    if (is.null(s) || !(is.numeric(s) && length(s) == 1))
      stop("For glmnet fits, provide a numeric lambda in 's'.")
    co <- as.matrix(stats::coef(fit, s = s))
  } else if (inherits(fit, "lm")) {
    # nice-to-have: also works for plain lm()
    cf <- stats::coef(fit)
    out <- data.frame(feature = names(cf), coefficient = as.numeric(cf), check.names = FALSE)
    if (requireNamespace("tibble", quietly = TRUE)) return(tibble::as_tibble(out))
    return(out)
  } else {
    stop("fit must be a glmnet or cv.glmnet (or lm) model.")
  }

  out <- data.frame(
    feature     = rownames(co),
    coefficient = as.numeric(co[, 1]),
    check.names = FALSE
  )
  if (requireNamespace("tibble", quietly = TRUE)) tibble::as_tibble(out) else out
}

# Plot coefficient paths for glmnet / cv.glmnet
# - glmnet_fit: a cv.glmnet or glmnet object
# - data: the training data frame used to fit the model (for sds)
# - features_to_plot: number of top features (by |coef| at s) to color and label
# - s: "lambda.1se" (default for cv.glmnet), "lambda.min", or a numeric lambda
plot_glmnet <- function(glmnet_fit, data, features_to_plot = 0,
                        s = "lambda.1se",
                        highlight = c("auto", "topk", "nonzero", "none"),
                        legend_nrow = "auto",       # set to a number to force rows
                        legend_max_cols = 3) {      # ~items per row when auto
  highlight <- match.arg(highlight)

  if (!requireNamespace("ggplot2", quietly = TRUE) ||
      !requireNamespace("scales", quietly = TRUE)) {
    stop("Please install 'ggplot2' and 'scales'.")
  }

  # -- pull fit and lambdas
  if (inherits(glmnet_fit, "cv.glmnet")) {
    fit <- glmnet_fit$glmnet.fit
    lambda_seq <- as.numeric(fit$lambda)
    s_val <- if (is.character(s)) {
      switch(s,
             "lambda.1se" = glmnet_fit$lambda.1se,
             "lambda.min" = glmnet_fit$lambda.min,
             stop("s must be 'lambda.1se', 'lambda.min', or numeric."))
    } else as.numeric(s)
  } else if (inherits(glmnet_fit, "glmnet")) {
    fit <- glmnet_fit
    lambda_seq <- as.numeric(fit$lambda)
    if (is.character(s)) stop("For glmnet fits, 's' must be numeric.")
    s_val <- as.numeric(s)
  } else stop("glmnet_fit must be a cv.glmnet or glmnet object.")

  # -- coefficients across lambda, drop intercept
  co <- as.matrix(stats::coef(fit, s = lambda_seq))
  feat_names <- rownames(co)[-1]
  co <- co[-1, , drop = FALSE]

  # -- standardize by predictor SDs
  sds <- vapply(feat_names, function(nm) stats::sd(data[[nm]], na.rm = TRUE), numeric(1))
  sds[is.na(sds)] <- 1
  co_std <- sweep(co, 1, sds, "*")

  # -- long data
  df <- data.frame(
    lambda      = rep(lambda_seq, each = nrow(co_std)),
    feature     = rep(feat_names, times = ncol(co_std)),
    coefficient = as.vector(co_std),
    stringsAsFactors = FALSE
  )

  # -- which features to color
  tol <- 1e-12
  co_at_s_std <- as.numeric(as.matrix(stats::coef(fit, s = s_val))[-1]) * sds
  nonzero_feats <- feat_names[abs(co_at_s_std) > tol]

  if (features_to_plot > 0 || highlight == "topk") {
    k <- max(1, min(features_to_plot, length(feat_names)))
    top_idx <- order(abs(co_at_s_std), decreasing = TRUE)[seq_len(k)]
    color_feats <- feat_names[top_idx]
  } else if (highlight == "nonzero") {
    color_feats <- nonzero_feats
  } else if (highlight == "auto") {
    color_feats <- if (length(nonzero_feats) > 0 &&
                       length(nonzero_feats) <= 0.5 * length(feat_names))
      nonzero_feats else character(0)
  } else {
    color_feats <- character(0)
  }

  # -- reversed log10 axis with nice breaks
  revlog10 <- scales::trans_new(
    "revlog10",
    transform = function(x) -log10(x),
    inverse   = function(x) 10^(-x),
    domain    = c(1e-300, Inf)
  )
  nice_lambda_breaks <- function(x) {
    rng  <- range(x, finite = TRUE)
    krng <- range(floor(log10(rng)), ceiling(log10(rng)))
    k    <- seq(krng[1], krng[2])
    cand <- c(1, 3) * rep(10^k, each = 2)
    cand[cand >= rng[1] & cand <= rng[2]]
  }

  library(ggplot2)
  p <- ggplot(df, aes(lambda, coefficient, group = feature)) +
    geom_line(color = "grey70", linewidth = 0.4) +
    geom_vline(xintercept = s_val, linetype = "dashed") +
    scale_x_continuous(
      trans  = revlog10,
      breaks = nice_lambda_breaks(lambda_seq),
      labels = scales::label_scientific(digits = 0),
      expand = expansion(mult = c(0.02, 0.02))
    ) +
    labs(x = expression(lambda), y = "Standardized Coefficient") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor.x = element_blank())
if (length(color_feats) > 0) {
    df_top <- df[df$feature %in% color_feats, ]
    n_items <- length(unique(df_top$feature))
    nrow_legend <- ceiling(n_items / 3)
    ncol_legend <- 3
    
    p <- p +
      geom_line(data = df_top, aes(color = feature), linewidth = 0.9) +
      # use scale_color_discrete(... guide = guide_legend(...)) to enforce nrow/ncol
      scale_color_discrete(
        guide = guide_legend(
          title = NULL,
          nrow = nrow_legend,
          ncol = ncol_legend,
          byrow = TRUE,
          override.aes = list(linewidth = 1.0)
        )
      ) +
      theme(
        legend.position  = "bottom",
        legend.direction = "horizontal",
        # this helps some themes/versions honor multi-row legends at the bottom
        legend.box       = "vertical"
      )
}
  p
}
```
# Instructions {.unnumbered}

## Materials and collaboration {.unlisted .unnumbered}

The policy on allowed materials and collaboration is as stated on the Syllabus:

>"Students are permitted to work together on homework assignments, but must write up and submit solutions individually. In particular, students may not copy each others’ solutions. Students may consult all course materials, textbooks, the internet, or AI tools (e.g. ChatGPT) to complete their homework. Students may not use solutions to problems that may be available online and/or from past iterations of the course. For each homework, students must disclose all classmates with whom they collaborated, which AI tools they used, and how they used them. Failure to do so will result in a 10-point penalty."

In accordance with this policy, 

*Please disclose all classmates with whom you collaborated:* None 

*Please disclose which AI tools you used, and how you used them:* ChatGPT in debugging r code and creating visualizations.

\textbf{\textcolor{red}{Failure to answer the above questions will result in a 10-point penalty.}}

## Writeup {.unlisted .unnumbered}

\textcolor{blue}{Use} `homework-3-template.Rmd` \textcolor{blue}{as a starting point for your writeup, adding your solutions after "**Solution**". Add your R code using code chunks and add your text answers using **bold text**.} Consult the [preparing reports guide](https://katsevich-teaching.github.io/stat-4710-fall-2023/assets/preparing-reports.pdf) for guidance on compilation, creation of figures and tables, and presentation quality. If the instructions ask you to "print a tibble", you should print the tibble directly. 


## Programming {.unlisted .unnumbered}

We encourage you to follow the `tidyverse` paradigm for data visualization, manipulation, and wrangling, as in the first homework, to practice tidyverse workflows (e.g., `filter`, `mutate`, `group_by`, `summarize`, `join`, `ggplot2`).

Students who strongly prefer to use Python for homework may do so. If you choose this option, you are **required** to present both code chunks and the corresponding outputs/interpretations in your PDF. In addition, it's recommended that you:

- Use `pandas` for data manipulation (preferred, since it parallels `dplyr`).
- Use `plotnine` for visualization (preferred, since it parallels ggplot2). If using seaborn/matplotlib, you must still include the required plot elements (labels, 45° line, smoothing curve, facets, etc.).  
 

```{r, message=FALSE}
library(tidyverse)
library(splines) 
library(glmnetUtils)
```

**Code completion.** In some problems, the task is to complete an incomplete code chunk provided. In those problems, in the template, there is `eval = FALSE` in the chunk header to prevent evaluation errors. You should remove `eval = FALSE` after you have completed the code to enable evaluation and printing of results.

## Grading {.unlisted .unnumbered}
The point value for each problem sub-part is indicated. There are 100 points possible on this homework, evaluated based on the correctness of the numerical results and plots. When a problem asks for a plot:

- **Axes and labels**: Always include informative axis labels and, if appropriate, titles.
- **Required elements**: If the instructions ask for a 45° line, smoothing curve, or facets, these must be included.
- **Clarity**: Points may be deducted if plots are cluttered, unreadable, or missing labels.

## Submission {.unlisted .unnumbered}

Compile your writeup to PDF and submit to Gradescope.

\newpage

# Case study: logistic regression for home loan (45 points)
 

## Data import (4 points)

1. **Solution.**
```{r}
hmeq = read_csv("home_loan.csv")
hmeq = drop_na(hmeq)
print(hmeq)
```
2. **Solution.**
```{r}
hmeq = hmeq |> mutate(STATUS = ifelse(STATUS == "bad", 1, 0))
hmeq$STATUS = factor(hmeq$STATUS, levels = c(0,1))
print(hmeq)
```

## Data exploration and visualization (4 points)

1. **Solution.**
```{r}
ggplot(data = hmeq, mapping = aes(x=DEBTINC)) + 
  geom_histogram(binwidth = 5) + 
  geom_vline(aes(xintercept = median(DEBTINC)), color = "red", linetype = "dashed") + 
  labs(title = "Distribution of DEBTINC with Median Line",
       x = "Debt-to-Income Ratio",
       y = "Count")
```
**Answer**: Based on the histogram, we noticed it's right skewed, where there are some outliers with heavy debt to income ratio. 

2. **Solution.**
```{r}
ggplot(hmeq, aes(x = STATUS, y = DEBTINC, fill = STATUS)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Debt-to-Income by Credit Status",
       x = "Credit Status",
       y = "Debt-to-Income Ratio") +
  theme_minimal()
```
**Answer**: we notice the paid loans generally have low debt to income ratio and default loans generally have outliers with high debt to income ratio. 

## Univariate logistic regression (23 points)
 

### Understanding logistic regression (13 points)

1. **Solution.**
```{r}
set.seed(4710) # seed set for reproducibility (DO NOT CHANGE) 
n <- nrow(hmeq)
train_samples <- sample(1:n, round(0.8 * n))
hmeq_train = hmeq[train_samples,]
hmeq_test = hmeq[-train_samples,]
```

2. **Solution.**
```{r}
hmeq_train_subset = hmeq_train |> 
  filter(VALUE < 50000 & YOJ == 0 & REASON == "DebtCon" & MORTDUE > 23000) |>
  select(STATUS, DEBTINC)
print(hmeq_train_subset)
```
3. **Solution.**
In a logistic regression model, the odds of default are given by

$$
\text{odds(default)} = \exp(\beta_0 + \beta_1 \times \text{DEBTINC})
$$

Therefore, the probability of default is

$$
p(\text{default}) = \frac{\exp(\beta_0 + \beta_1 \times \text{DEBTINC})}
{1 + \exp(\beta_0 + \beta_1 \times \text{DEBTINC})}.
$$

Given four observations, with two defaulted and two paid loans, the likelihood function is

$$
\begin{aligned}
L(\beta_0, \beta_1)
&= \prod_{i=1}^{4} p_i^{y_i} [1 - p_i]^{1 - y_i} \\[6pt]
&= 
\left[\frac{\exp(\beta_0 + \beta_1 \times 25.3)}
{1 + \exp(\beta_0 + \beta_1 \times 25.3)}\right]  % y=1
\times
\left[1 - \frac{\exp(\beta_0 + \beta_1 \times 16.5)}
{1 + \exp(\beta_0 + \beta_1 \times 16.5)}\right]  % y=0
\times \\[8pt]
&\quad
\left[\frac{\exp(\beta_0 + \beta_1 \times 29.3)}
{1 + \exp(\beta_0 + \beta_1 \times 29.3)}\right]  % y=1
\times
\left[1 - \frac{\exp(\beta_0 + \beta_1 \times 16.7)}
{1 + \exp(\beta_0 + \beta_1 \times 16.7)}\right].  % y=0
\end{aligned}
$$
4. **Solution.**
```{r}
# fit a logistic model based on the subset training data 
fit1 = glm(STATUS~DEBTINC, family="binomial", data=hmeq_train_subset)

# identify beta_0 be -115.38, beta_1 be 5.47
summary(fit1)
```
**Answer:**  
The estimated probability of default, given a debt-to-income ratio \(x\), is

$$
P(\text{STATUS} = 1 \mid \text{DEBTINC} = x)
= 
\frac{\exp(-115.38 + 5.47x)}
     {1 + \exp(-115.38 + 5.47x)}.
$$
5. **Solution.**
Fitted values of $\beta_0$ and $\beta_1$ are obtained by maximizing the likelihood (and log likelihood) function formulated in part 3. 

### Univariate logistic regression on the full data (10 points)

1. **Solution.**
```{r}
# fit a logistic model using the whole training dataset
fit_2 = glm(STATUS~DEBTINC, family = "binomial", data=hmeq_train)
summary(fit_2)
```
2. **Solution.**
Based on the model summary above, we observe for 1 increase in DEBTINC, the odds
of default will multiply by exp(0.0911) (1.095). 

3. **Solution.**
```{r}
ggplot(hmeq_train, aes(x = DEBTINC, y = 1 * (STATUS == 1))) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.5, color = "darkblue") +
  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, color = "red", size = 1) +
  labs(
    x = "Debt-to-Income Ratio (DEBTINC)",
    y = "Probability of Default (STATUS = 1)",
    title = "Probability of Default vs. Debt-to-Income"
  ) +
  theme_minimal()
```
**Answer**: Based on the plot, when DEBTINC is 50, the estimated probability of default is roughly 0.4. 

## Multivariate logistic regression (14 points)

1. **Solution.**
```{r}
fit_3 = glm(STATUS~., family = "binomial", data = hmeq_train)
summary(fit_3)
```
**Answer**: Based on the summary, we observe a higher DEBTINC are more likely to default when other things are held equal. And other variables are held equal, for 1 increase in NINQ, it is associated with exp(0.154) = 1.17 times more likely in terms of odds of defaults. 

2. **Solution.** 
```{r}
# create a dataframe for the new data
new_data <- data.frame(
  LOAN = 10000,
  MORTDUE = 65000,
  VALUE = 90000,
  REASON = "DebtCon",
  JOB = "Sales",
  YOJ = 5,
  DEROG = 0,
  DELINQ = 0,
  CLAGE = 120,
  NINQ = 1,
  CLNO = 15,
  DEBTINC = 50
)

# predict the new data point 
predicted_prob <- predict(fit_3, newdata = new_data, type = "response")

print(predicted_prob)
```
**Answer**: There is roughly 71.2% chance of defualt given above features. 

3. **Solution.**
```{r}
# obtain predicted prob
prob_hat = predict(fit_3, newdata = hmeq_test, type = "response")

# classify using a 0.5 threshold 
pred_class = ifelse(prob_hat >= 0.5, "1", "0")
pred_class = factor(pred_class, levels = c("0","1"))

# calculate four components in confusion matrix
tp = sum(pred_class == "1" & hmeq_test$STATUS == "1")
tn = sum(pred_class == "0" & hmeq_test$STATUS == "0")
fp = sum(pred_class == "1" & hmeq_test$STATUS == "0")
fn = sum(pred_class == "0" & hmeq_test$STATUS == "1")

# calculate metrics
n  = nrow(hmeq_test)
misclass_rate = (fp + fn) / n
precision = tp / (tp + fp)
recall = tp / (tp + fn)
F1 = 2 * precision * recall / (precision+ recall)
  
test_errors <- tribble(
  ~Metric, ~Value, 
  "Misclassification_rate", misclass_rate,
  "Precision", precision,
  "Recall", recall,
  "F1_score", F1
)
# print the tibble
print(test_errors)
```


\newpage

# Case study: linear regression for housing prices (55 points)
 

## Exploratory data analysis (10 points)

1. **Solution.**
```{r}
# read in data
ames_data = read_csv("ames_housing.csv")
print(ames_data)
```
2. **Solution.**
```{r}
# visualize the log price with 
ggplot(data = ames_data, aes(x = log_Sale_Price)) + 
  geom_histogram(binwidth = 0.1) + 
  geom_vline(aes(xintercept = median(log_Sale_Price), color = "red")) + 
  labs(title = "log sales price histogram")
```
**Answer**: The distribution of log sales price is roughly symmetric with some outliers at the lower end. 

3. **Solution.**
```{r}
# Scatter plot 1: log_Sale_Price vs. Gr_Liv_Area
ggplot(ames_data, aes(x = Gr_Liv_Area, y = log_Sale_Price)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(
    title = "Log Sale Price vs. Above-Ground Living Area",
    x = "Above-Ground Living Area (sq ft)",
    y = "Log of Sale Price"
  ) +
  theme_minimal()

# Scatter plot 2: log_Sale_Price vs. Year_Built
ggplot(ames_data, aes(x = Year_Built, y = log_Sale_Price)) +
  geom_point(alpha = 0.2, color = "darkgreen") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(
    title = "Log Sale Price vs. Year Built",
    x = "Year Built",
    y = "Log of Sale Price"
  ) +
  theme_minimal()
```
**Answer**: Based on the two plot above, we observe the log sale price tends to increase with respect to above ground living area before 4000 sq ft then decrease. And the log sales price tends to decrease when year build before 1940s then increase after that. 

## Predictive modeling preparation (1 point)
 

**Solution.**

```{r}
set.seed(4710) # seed set for reproducibility (DO NOT CHANGE)
n <- nrow(ames_data)
train_samples <- sample(1:n, round(0.8 * n))
ames_train = ames_data[train_samples,]
ames_test = ames_data[-train_samples,]
```
 

## Ordinary least squares (6 points)

1. **Solution.**
```{r}
ols_fit = lm(log_Sale_Price ~ . ,data=ames_train)
summary(ols_fit)
```
2. **Solution.**
Some NA values are in the fitted coefficients because some features are linearly dependent with others. For instance, some categorical variables, like sale condition, are not entered as factors (of which R will automatically drop one category), hence resulting the last category lienarly dependent on other categories in the same variable. 

3. **Solution.**
Despite both signs of Gr_Liv_Area and Year_Built with respect to log Sale Price are positive, those two values are close to zero (2.66e-3 for Year Build and 1.45e-4 for Gr_live_area), which does not match the patterns we see in previous two plots. 


## Ridge regression (13 points)

1. **Solution.**
```{r}
set.seed(471)
ridge_fit <- cv.glmnet(
  log_Sale_Price ~ ., 
  alpha = 0,                # alpha = 0 for ridge (stay tuned for more)
  nfolds = 10,              # number of folds
  data = ames_train        # data to run ridge on 
)

# plot the cv mean error as a function of lambda
plot(ridge_fit)

# identify 1se for lambda
ridge_fit$lambda.1se
```
**Answer**: The lambda for 1 SE rule is 0.672. 

2. **Solution.**
**Answer**: Since each fold we have 100 fit on for each value of lambda, in total we have 10 * 100 = 1000 model fit excluding the final fit on the whole data. 

3. **Solution.**
**Answer**: Since each data is used in 9 training fold and 1 validation fold, each data is used 9 * 100 = 900 times using it. 

4. **Solution.**
```{r}
# visualize ridge coefs 
plot_glmnet(ridge_fit, ames_train, features_to_plot = 6)
```
**Answer**: None of the 6 features change sign as lambda decrease and only some of the 6 features change monotonically as lambda decreases. 


## Lasso regression (10 points)

1. **Solution.**
```{r}
set.seed(471)
lasso_fit <- cv.glmnet(
  log_Sale_Price ~ ., 
  alpha = 1,                # alpha = 1 for lasso 
  nfolds = 10,              # num of folds
  data = ames_train        # data to run on 
)

# plot the cv mean error as a function of lambda
plot(lasso_fit)
```
2. **Solution.**
```{r}
# check 1 se lambda 
print(lasso_fit$lambda.1se)

# extract 1se lambda coefs
beta_1se = as.matrix(coef(lasso_fit, s = lasso_fit$lambda.1se))

# remove the intercept
beta_no_intercept = beta_1se[-1, , drop = FALSE]

# identify nonzero and zero coefficients
nz_idx = which(beta_no_intercept[, 1] != 0)
z_idx = which(beta_no_intercept[, 1] == 0)

# Compute counts
num_nonzero = length(nz_idx)
print(num_nonzero)
```
**Answer**: There are 27 features have nonzero coefficients in the Lasso fit with 1 se lambda, which is 0.023. 

3. **Solution.**
```{r}
# visualize lasso coefs 
plot_glmnet(lasso_fit, ames_train, features_to_plot = 8)
```
**Answer**: Based on the plot, Gr_Liv_Area is the first feature to enter the model as lambda decreases and the same feature (Gr_Liv_Area) has the largest absolute coefficient for the most flexibly fitted Lasso. 

## Evaluation and comparison (15 points)

1. **Solution.**

```{r}
# get three models predictions
ols_predictions = predict(ols_fit, newdata = ames_test) |> as.numeric()

ridge_predictions = predict(ridge_fit,
  newdata = ames_test,
  s = "lambda.1se"
) |> as.numeric()

lasso_predictions = predict(lasso_fit,
  newdata = ames_test,
  s = "lambda.1se"
) |> as.numeric()

# extract y label
y_test = ames_test$log_Sale_Price

# compute mse
mse_ols = mean((y_test - ols_predictions)^2)
mse_ridge = mean((y_test - ridge_predictions)^2)
mse_lasso = mean((y_test - lasso_predictions)^2)

error_for_models <- tribble(
  ~Model, ~RMSE, 
  "Linear", mse_ols, 
  "Ridge", mse_ridge,
  "Lasso", mse_lasso
)
# print the tibble
print(error_for_models)
```

2. **Solution.**
**Answer** Based on the above result, we observe the OLS has the lowest RMSE compared with Ridge and Lasso regression. Since regularization models like Ridge and Lasso are used to decrease variance and increase bias, this indicates bias is the dominant driving force for test error. This makes sense, since our whole dataset has 2930 data points, which is fairly large dataset for 172 features. 

3. **Solution.**
```{r}
# extracting coefs
ls_coef = coef(ols_fit) |> as.numeric()
ridge_coef = coef(ridge_fit, s = "lambda.1se") |> as.matrix()
lasso_coef = coef(lasso_fit, s = "lambda.1se") |> as.matrix()

# get feature names
feature_names = rownames(ridge_coef)[-1]  
ls_coef_values = ls_coef[-1]
ridge_coef_values = ridge_coef[-1, 1]
lasso_coef_values = lasso_coef[-1, 1]

# create a coef tibble 
coeffs = tibble(
  feature = feature_names,
  ls_coef = ls_coef_values,
  ridge_coef = ridge_coef_values,
  lasso_coef = lasso_coef_values
) |>
  arrange(desc(abs(ls_coef)))

# print tibble
print(coeffs)
```
**Answer**: we notice the coefficients magnitude from OLS is always larger than Ridge model, since OLS is the case for lambda be zero for Ridge model. 

4. **Solution.**
```{r}
# get rows for ls coef be na 
na_rows = coeffs |> filter(is.na(ls_coef))
print(na_rows)
```
**Answer**: For the OLS NA fitted coefficients, the ridge and lasso models do not have them be NA. Ridge addresses the issue by assigning dependent features similar coefficients values and Lasso addresses the issue by randomly assigning one of the dependent features 0 coefficients. 