---
title: 'Unit 2 Lecture 3: Cross-validation'
date: "September 16, 2025"
output: pdf_document
---

In this R demo, we will implement cross-validation to select the degrees of freedom of a natural spline fit, using the running example from the previous class. We'll need the following R packages:
```{r, message = FALSE}
library(tidyverse)
library(splines)   # for ns()
library(cowplot)   # for plot_grid()
theme_set(
  theme_bw()  
) 
```

```{r, echo = FALSE}
cross_validate_spline <- function(x, y, nfolds = 10, df_values = 2:15, seed = NULL) {
  stopifnot(length(x) == length(y))
  n <- length(x)
  if (!is.null(seed)) set.seed(seed)

  # Same fold construction style as in the Rmd
  folds <- sample(rep(1:nfolds, length.out = n))
  train_df <- data.frame(x = x, y = y, fold = folds)

  # Helper: compute CV mean & SE for a single displayed df value
  # NOTE: to match your validation plot, we fit ns(x, df = df_display - 1)
  cv_for_df <- function(df_display) {
    df_use <- df_display - 1
    oof <- numeric(n)
    for (k in 1:nfolds) {
      out <- train_df[train_df$fold != k, , drop = FALSE]
      in_  <- train_df[train_df$fold == k, , drop = FALSE]
      fit <- lm(y ~ ns(x, df = df_use), data = out)
      oof[folds == k] <- predict(fit, newdata = in_)
    }
    # fold-wise CV estimates (same aggregation as in the Rmd)
    cv_by_fold <- tapply((oof - train_df$y)^2, INDEX = train_df$fold, FUN = mean)
    c(mean = mean(cv_by_fold), se = sd(cv_by_fold) / sqrt(nfolds))
  }

  # Run across all df values
  mat <- t(vapply(df_values, cv_for_df, numeric(2)))
  cv_table <- data.frame(
    df = df_values,
    cv_mean = mat[, "mean"],
    cv_se   = mat[, "se"]
  )

  # df.min and 1SE rule (prefer simpler/smaller df within 1 SE)
  min_row <- cv_table[which.min(cv_table$cv_mean), ]
  df.min  <- min_row$df
  thresh  <- min_row$cv_mean + min_row$cv_se
  df.1se  <- with(cv_table[cv_table$cv_mean <= thresh, ], min(df))

  # Plot (points + line + error bars; dashed lines at min and 1SE)
  cv_plot <- ggplot(cv_table, aes(df, cv_mean)) +
    geom_line(linewidth = 0.8, color = "black") +
    geom_point(size = 2, color = "black") +
    geom_errorbar(aes(ymin = cv_mean - cv_se, ymax = cv_mean + cv_se),
                  width = 0.3, color = "black") +
    geom_hline(yintercept = min_row$cv_mean, linetype = 2) +
    geom_hline(yintercept = thresh, linetype = 2) +
    geom_vline(xintercept = df.min, linetype = 3) +
    geom_vline(xintercept = df.1se, linetype = 3) +
    labs(x = "Degrees of freedom", y = "CV error") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  list(cv_table = cv_table, cv_plot = cv_plot, df.1se = df.1se, df.min = df.min)
}
```

# Training and validation

Let us create a training set:
```{r}
set.seed(1)
f <- function(x) (sin(3 * x))
n <- 50
sigma <- 1
train_data <- tibble(
  x = seq(0, 2 * pi, length.out = n),
  y = f(x) + rnorm(n, sd = sigma)
)
train_data
```

Let's also suppose we have a large validation set on our hands:
```{r}
N <- 50000
validation_data <- tibble(
  x = seq(0, 2 * pi, length.out = N),
  y = f(x) + rnorm(n, sd = sigma)
)
```

Now let's fit splines with `df` = 2,3,...,15 to the training data, and evaluate their test error using the test set:

```{r, fig.width = 4, fig.height = 3}
# compute the validation error
max_df <- 15
validation_error <- numeric(max_df-1)
for (df in 2:max_df) {
  formula <- sprintf("y ~ ns(x, df = %d)", df - 1)
  spline_fit <- lm(formula = formula, data = train_data)
  y_hat_validation <- predict(spline_fit, newdata = validation_data)
  validation_error[df - 1] <- validation_data |>
    mutate(y_hat_validation = y_hat_validation) |>
    summarize(mean((y_hat_validation - y)^2)) |>
    pull()
}

# plot the validation error
p_val <- tibble(df = 2:max_df, validation_error) |>
  ggplot(aes(x = df, y = validation_error)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Degrees of freedom",
    y = "Validation error"
  )
plot(p_val)
```

The issue is that we usually do not have a giant validation set for model selection purposes. We need to make do with our smallish training set for both model training and model selection. This is where cross-validation comes in handy!

# Cross-validation for df = 5

The idea is to split our training samples into *folds* and then have the folds take turns being the validation set. Let's take a look.

```{r}
K <- 10
folds <- sample(rep(1:K, n / K))
train_data <- train_data |>
  mutate(fold = folds)
train_data
```
Question: How would we select the data in fold number 1? How would we select all the data except fold number 1?

Let's first use cross-validation to estimate the test error for a spline fit with 5 degrees of freedom.
```{r}
# create a vector of out-of-fold predictions
out_of_fold_predictions <- numeric(n)

# iterate over folds
for (current_fold in 1:K) {
  # out-of-fold data will be used for training
  out_of_fold_data <- train_data |> filter(fold != current_fold)
  # in-fold data will be used for validation
  in_fold_data <- train_data |> filter(fold == current_fold)

  # train on out-of-fold data
  spline_fit <- lm(y ~ ns(x, df = 5), data = out_of_fold_data)

  # predict on in-fold data
  out_of_fold_predictions[folds == current_fold] <-
    predict(spline_fit, newdata = in_fold_data)
}

# add the out-of-fold predictions to the data frame
results <- train_data |>
  mutate(yhat = out_of_fold_predictions)
results

# compute the CV estimate and standard error
results |>
  summarize(cv_fold = mean((yhat - y)^2), .by = fold) |> # CV estimates per fold
  summarize(
    cv_mean = mean(cv_fold),
    cv_se = sd(cv_fold) / sqrt(K)
  )
```
What are two reasons this CV estimate may be different from the validation error estimated above?

# Cross-validation for df = 2,3,...,15

Now let's repeat what we did above for many degrees of freedom, because after all, the point of cross-validation is to choose the degrees of freedom. We can do this via `cross_validate_spline()`. 

```{r}
cv_results <- cross_validate_spline(
  x = train_data$x,
  y = train_data$y,
  nfolds = 10,
  df_values = 2:15
)
```

The `cv_results` variable stores an *object*, with the following fields:
```{r}
names(cv_results)
```

Let's inspect these one at a time:
```{r}
cv_results$cv_table
```

```{r, fig.width = 4, fig.height = 3}
cv_results$cv_plot
```

Let's compare the CV plot to the validation error plot:
```{r, fig.width = 4, fig.height = 6}
plot_grid(cv_results$cv_plot, p_val, nrow = 2)
```

Based on this plot, what degrees of freedom would we select using the one-standard error rule?

```{r}
cv_results$df.1se
```

```{r}
cv_results$df.min
```

# Exercise

Use 10-fold cross-validation and the one-standard error rule to select the optimal number of degrees of freedom for regressing `income` on `age` in the `income_2007` data from Unit 3 Lecture 1. Make the CV plot, extract `df.1se` and produce a scatter plot of `income` versus `age` with this optimal spline fit superimposed.
```{r, message = FALSE}
income_2007 <- read_csv("income_data.csv") |> 
  filter(year == 2007) |> 
  select(-year)
```
