---
title: 'Unit 5 Lecture 3: Convolutional Neural Networks'
date: "November 13, 2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

In this R demo, we'll be fitting convolutional neural networks to the MNIST handwritten digit data. A python version is here: https://colab.research.google.com/drive/1SRRXmXtkgCMJNJu-GLcJfpv6F2WWq7O6?usp=sharing

First let's load some libraries:
```{r, message = FALSE} 
library(cowplot)   # for side-by-side plots 
library(tidyverse)  
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot theme
options(warn=-1) # optional, to suppress warnings; comment out if you want to see warnings 
library(torch)
library(luz)
library(torchvision)
```


Next let's load the MNIST data as before. 
```{r}
# MNIST as tensors in [0,1]
train_full <- mnist_dataset(root = "../data", train = TRUE, download = TRUE,
                            transform = transform_to_tensor)
test_ds    <- mnist_dataset(root = "../data", train = FALSE, download = TRUE,
                            transform = transform_to_tensor)

# basic sizes
img_rows <- 28; img_cols <- 28; P <- img_rows * img_cols; K <- 10

# 20% validation split (like validation_split=0.2)
n <- length(train_full)
v <- floor(0.2 * n)
valid_ds <- dataset_subset(train_full, 1:v)
train_ds <- dataset_subset(train_full, (v+1):n)

# dataloaders (SGD/mini-batches per your slides)  :contentReference[oaicite:1]{index=1}
train_dl <- dataloader(train_ds,  batch_size = 128, shuffle = TRUE)
valid_dl <- dataloader(valid_ds,  batch_size = 128, shuffle = FALSE)
test_dl  <- dataloader(test_ds,   batch_size = 256, shuffle = FALSE)
```

# Build convolutional neural network 

Next, we define a convolutional neural network model with two convolutional layers, one max pooling layer, and one dense layer.
```{r}
# model: Conv(32,3x3)->ReLU -> Conv(64,3x3)->ReLU -> MaxPool(2) -> Dropout(0.25)
#        -> Flatten -> Dense(128)->ReLU -> Dropout(0.5) -> Dense(10)
cnn <- nn_module(
  "SmallCNN",
  initialize = function() { # define model structure
    self$conv1 <- nn_conv2d(in_channels=1, out_channels=32, kernel_size=3)
    self$conv2 <- nn_conv2d(32, 64, kernel_size=3)
    self$pool  <- nn_max_pool2d(kernel_size=2)
    self$drop1 <- nn_dropout(0.25)
    self$flatten <- nn_flatten()
    self$fc1   <- nn_linear(64*12*12, 128)  # after 2 conv (no padding) + 2x2 pool on 28x28
    self$drop2 <- nn_dropout(0.5)
    self$fc2   <- nn_linear(128, 10)
  },
  forward = function(x) { # define forward pass
    x <- nnf_relu(self$conv1(x))
    x <- nnf_relu(self$conv2(x))
    x <- self$pool(x)
    x <- self$drop1(x)
    x <- self$flatten(x)   # flatten [B,64,12,12] -> [B,9216]
    x <- nnf_relu(self$fc1(x))
    x <- self$drop2(x)
    self$fc2(x)                           # logits; CE applies softmax internally
  }
)
```
 

# Train the CNN with adam

To train this neural network, we must first define what loss function to use, which optimizer to use, and which metrics to track. We do this by *compiling* the model.
```{r}
dir.create("checkpoints", showWarnings = FALSE)
fitted <- cnn %>%
  setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_adam,
    metrics = list(luz_metric_accuracy())
  ) %>%
  set_opt_hparams(lr = 1e-3) %>%
  fit(
    train_dl, epochs = 3, valid_data = valid_dl,
    callbacks = list(
      luz_callback_early_stopping(patience = 2, monitor = "valid_loss") 
    )
  )
```

Let's plot the training/validation curves:

```{r}
plot(fitted)          # training/validation curves
```

# Evaluate on test data

We can evaluate the accuracy on test data as before:

```{r}
evaluate(fitted, test_dl)
```

We can also manually create the predictions:

```{r}
# predictions
logits <- predict(fitted, test_dl)
probs  <- nnf_softmax(logits, dim=2)
pred_i <- as.integer(as.array(torch_argmax(probs, dim=2)))  # 1..10
pred   <- pred_i - 1                                        # 0..9 for display
# true labels (1..10) -> digits 0..9
y_true_i <- c(); for (b in enumerate(test_dl)) y_true_i <- c(y_true_i, as.integer(as.array(b[[2]])))
y_true <- y_true_i - 1

mean(pred == y_true)    # accuracy
```