---
title: 'Unit 5 Lecture 2: Neural Networks'
date: "November 11, 2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

In this R demo, we'll be fitting fully-connected neural networks to the MNIST handwritten digit data. Its python version is here: https://colab.research.google.com/drive/1Qo3VeziJ24lkxcvM9KEt9WhYmIJsBCwX?usp=sharing 

First let's load some libraries:
```{r, message = FALSE} 
library(cowplot)   # for side-by-side plots 
library(tidyverse)  
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot theme
options(warn=-1) # optional, to suppress warnings; comment out if you want to see warnings 
library(torch)
library(luz)
library(torchvision)
```

Next let's load the MNIST data as before. 
```{r}
# MNIST as tensors in [0,1]
train_full <- mnist_dataset(root = "../data", train = TRUE, download = TRUE,
                            transform = transform_to_tensor)
test_ds    <- mnist_dataset(root = "../data", train = FALSE, download = TRUE,
                            transform = transform_to_tensor)

# basic sizes
img_rows <- 28; img_cols <- 28; P <- img_rows * img_cols; K <- 10

# 20% validation split (like validation_split=0.2)
n <- length(train_full)
v <- floor(0.2 * n)
valid_ds <- dataset_subset(train_full, 1:v)
train_ds <- dataset_subset(train_full, (v+1):n)

# dataloaders (SGD/mini-batches per your slides)  :contentReference[oaicite:1]{index=1}
train_dl <- dataloader(train_ds,  batch_size = 128, shuffle = TRUE)
valid_dl <- dataloader(valid_ds,  batch_size = 128, shuffle = FALSE)
test_dl  <- dataloader(test_ds,   batch_size = 256, shuffle = FALSE)
```

# Model: 1 hidden layer (256) + ReLU + Dropout + Softmax-at-loss

Next, we define a neural network model with one hidden layer with 256 units and dropout rate 0.5. 

```{r}
mlp <- nn_module(
  "MLP256",
  initialize = function() {
    self$flatten <- nn_flatten()
    self$fc1  <- nn_linear(P, 256)
    self$drop <- nn_dropout(0.5)
    self$fc2  <- nn_linear(256, K)
  },
  forward = function(x) {
    x <- self$flatten(x) # x$view(base::c(x$size(1), P))
    x <- nnf_relu(self$fc1(x))
    x <- self$drop(x)
    self$fc2(x)   # logits
  }
)

fitted <- mlp %>%
  setup(loss = nn_cross_entropy_loss(),
        optimizer = optim_adam,
        metrics = list(luz_metric_accuracy())) %>%
  set_opt_hparams(lr = 1e-2) %>%
  fit(train_dl, epochs = 5, valid_data = valid_dl,
      callbacks = list(
        luz_callback_early_stopping(patience = 3, monitor = "valid_loss")#,
        # luz_callback_model_checkpoint(path = "checkpoints/best.pt",
        #                               monitor = "valid_loss", save_best_only = TRUE)
      ))
print(fitted)
```


# Save & reload the trained model + history

```{r}
# Save (model + optimizer state, etc.)
luz_save(fitted, "mlp_256.pt") # can also save as "mlp_256.rds" and load accordingly

# Reload later
fitted <- luz_load("mlp_256.pt")

# History plot stays available via plot(fitted)
plot(fitted)
```

# Evaluate on test data

```{r, message=FALSE}
logits <- predict(fitted, test_dl)
probs  <- nnf_softmax(logits, dim = 2)
pred   <- as.integer(as.array(torch_argmax(probs, dim = 2)))

# True labels
y_true <- c()
for (b in enumerate(test_dl)){
  y_true <- c(y_true, as.integer(as.array(b[[2]])))
} 
mean(pred == y_true)                # accuracy
table(Predicted = pred, True = y_true) 
```
